---
title: "Inequality and Poverty in Brazil."
author: "Vitor Hugo Miro (UFC)"
date: "`r format(Sys.time(), '%d %B, %Y')`"

output: html_notebook

---

## Poverty

This analysis are based on data from the Continuous National Household Sample Survey (Pesquisa Nacional por Amostra de Domicílios Contínua) for 2019.

We will be using R and RStudio, as well as the appropriate libraries, to extract the data, calculate poverty and inequality indicators, and plot graphs to present this information.


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE, include=FALSE}
rm(list = ls())
options(scipen = 999)
```

```{r, include=TRUE}
# Carregar pacote
library(PNADcIBGE)
library(survey)
library(srvyr)
library(convey)
library(tidyverse)
```

```{r, include=TRUE}
# Definir variáveis utilizadas na análise em um vetor

variaveis <- c('Ano', 'UF', 'Estrato', 'UPA', 'V1008', 'V1014', 'V1032', 'V1022', 'V2009', 'VD2002', 'VD4019', 'VD4048')

```


```{r, include=TRUE}
# Carrega dados da PNADC 
pnadc <- get_pnadc(year = 2019,
                   vars = variaveis,
                   interview = 1,
                   design = TRUE,
                   labels = TRUE,
                   deflator = TRUE,
                   defyear = 2019)
```

```{r, echo = TRUE}
pnadc <- as_survey(pnadc)
```


```{r}
pnadc <- pnadc %>% 
    mutate(
        id_dom = as.numeric(paste(UPA,V1008,V1014, sep = "")),
        membro = ifelse(VD2002=="Pensionista" | 
                            VD2002=="Empregado(a) doméstico(a)" | 
                            VD2002=="Parente do(a) empregado(a) doméstico(a)",
                        0,1),
        ### Contruir variaveis de renda
        rend_trabalho = ifelse (is.na (VD4019), 0, VD4019*CO2),
        rend_outrasf = ifelse (is.na (VD4048), 0, VD4048*CO2e),
        renda = rend_trabalho + rend_outrasf,
        ### Dimensões geograficas
        brasil = factor("Brasil"),
        regiao = factor(ifelse(substr(UPA, start=1, stop=1)=="1","Norte",
                                  ifelse(substr(UPA, start=1, stop=1)=="2","Nordeste",
                                         ifelse(substr(UPA, start=1, stop=1)=="3","Sudeste",
                                                ifelse(substr(UPA, start=1, stop=1)=="4","Sul",
                                                       ifelse(substr(UPA, start=1, stop=1)=="5","Centro-Oeste",NA))))),
                        levels = c("Norte","Nordeste","Sudeste","Sul","Centro-Oeste"))
    )


# Criar variaveis agregadas por domicilio
pnadc <- pnadc %>% group_by(id_dom) %>% 
    mutate(n_dom = sum(membro),
           # Construir variavel de renda domiciliar total e rdpc
           rend_dom_outrasf = ifelse(membro==1, sum(rend_outrasf), 0),
           rend_dom_trabalho = ifelse(membro==1, sum(rend_trabalho), 0),
           renda_dom = ifelse(membro==1, sum(renda), 0),
           rdpc = renda_dom/n_dom ) %>%
    ungroup()
```

## Measuring poverty

The poverty measures, commonly known as the Foster-Greer-Thorbecke (FGT) family of poverty, including:

- Poverty headcount index = FGT(0)
- Poverty gap index = FGT(1)
- Poverty severity index = FGT(2)


This poverty measures was calculated using the R `convey` library.

```{r, echo = TRUE}
pnadc <- convey_prep(pnadc)
class(pnadc)
```
**Poverty lines**
The World Bank’s poverty line are widely used international poverty lines, including US\$1.9 a day, US\$3.2 a day and US\$ 5.5 a day (in 2011 PPP). 


```{r, echo = TRUE}
fgt0 <- svyfgt(~rdpc, pnadc, g=0, abs_thresh = 436, na.rm = TRUE)
fgt0 <- round(fgt0, 4)*100

fgt1 <- svyfgt(~rdpc, pnadc, g=1, abs_thresh = 436, na.rm = TRUE)
fgt1 <- round(fgt1, 4)*100

fgt2 <- svyfgt(~rdpc, pnadc, g=2, abs_thresh = 436, na.rm = TRUE)
fgt2 <- round(fgt2, 4)*100
```

Considering the World Bank poverty line (an income of up to US\$5.5 per day or R\$436 per month), the poverty indicators are as follows:

- Poverty headcount index = `r (fgt0)`%.

- Poverty gap index = `r (fgt1)`%.

- Poverty severity index = `r (fgt2)`%.


## Poverty rate in 2019, by state (Unidade Federativa - UF)

Brazil is a country with a large territorial dimension and its regions present very heterogeneous social indicators.

```{r, echo = TRUE}
fgt0_uf <- survey::svyby(~rdpc,
                         by = ~UF,
                         design = pnadc,
                         FUN = svyfgt,
                         g=0, 
                         abs_thresh = 436,
                         na.rm = TRUE)
```

The graph below shows the poverty rate values for the Brazilian states.

```{r}

Norte <- c("Acre", "Amapá", "Amazonas", "Pará", "Rondônia", "Roraima", "Tocantins")
Nordeste <- c("Alagoas", "Bahia", "Ceará", "Maranhão", "Paraíba", "Pernambuco", 
              "Piauí", "Rio Grande do Norte", "Sergipe")
COeste <- c("Distrito Federal", "Goiás", "Mato Grosso", "Mato Grosso do Sul")
Sudeste <- c("Espírito Santo", "Minas Gerais", "Rio de Janeiro", "São Paulo")
Sul <- c("Paraná", "Rio Grande do Sul", "Santa Catarina")


fgt0_uf <- fgt0_uf %>% mutate(
    regiao = factor(case_when(UF %in% Norte ~ "Norte",
                              UF %in% Nordeste ~ "Nordeste",
                              UF %in% COeste ~ "Centro-Oeste",
                              UF %in% Sudeste ~ "Sudeste",
                              UF %in% Sul ~ "Sul")),
    abbrev_uf = recode(UF, 
                       "Acre" = "AC",
                       "Amapá" = "AP",
                       "Amazonas" = "AM",
                       "Pará" = "PA",
                       "Rondônia" = "RO",
                       "Roraima" = "RR",
                       "Tocantins" = "TO",
                       "Alagoas" = "AL",
                       "Bahia" = "BA",
                       "Ceará" = "CE",
                       "Maranhão" = "MA",
                       "Paraíba" = "PB",
                       "Pernambuco" = "PE",
                       "Piauí" = "PI",
                       "Rio Grande do Norte" = "RN",
                       "Sergipe" = "SE",
                       "Distrito Federal" = "DF",
                       "Goiás" = "GO",
                       "Mato Grosso" = "MT", 
                       "Mato Grosso do Sul" = "MS",
                       "Espírito Santo" = "ES", 
                       "Minas Gerais" = "MG", 
                       "Rio de Janeiro" = "RJ", 
                       "São Paulo" = "SP",
                       "Paraná" = "PR", 
                       "Rio Grande do Sul" = "RS", 
                       "Santa Catarina"= "SC")
    )

fgt0_uf %>% 
    dplyr::as_tibble() %>%
    dplyr::mutate(UF = forcats::fct_reorder(UF, rdpc)) %>%
    ggplot2::ggplot(ggplot2::aes(x = rdpc, y = UF)) +
    ggplot2::geom_col(aes(fill = regiao)) +
    ggplot2::labs(title = "Poverty rate by UF (2019)",
                  subtitle = "Considering the US$ 5.50 PPP line.",
                  x = NULL,
                  y = NULL,
                  caption = "Source: IBGE/PNADC microdata.")+
    ggplot2::geom_text(aes(label= round(rdpc,2)), vjust=0.5, hjust=-0.2, size=2.5)

```

**Poverty Map**
```{r}
library(geobr)
library(sf)

map <- read_state(showProgress = FALSE)

fgt0_map = left_join(map, fgt0_uf, by=c("abbrev_state"="abbrev_uf"))

fgt0_map %>% 
    ggplot(aes(fill=rdpc), color = "black") +
    geom_sf() +
    theme_bw() +
    scale_fill_distiller(palette = "YlGnBu", name="Percentage", trans = "reverse")+
    labs(title="Percentage of People in Poverty by State",
         caption="Source: IBGE/PNADC microdata.", size=8)

```

### Poverty rate in rural areas

```{r, echo = TRUE}
fgt0_uf_rural <- survey::svyby(~VD5008,
                               by = ~UF,
                               design = subset(pnadc, V1022 == "Rural"),
                               FUN = svyfgt,
                               g=0, 
                               abs_thresh = 436,
                               na.rm = TRUE)

fgt0_uf_rural <-fgt0_uf_rural %>% mutate(
    regiao = factor(case_when(UF %in% Norte ~ "Norte",
                              UF %in% Nordeste ~ "Nordeste",
                              UF %in% COeste ~ "Centro-Oeste",
                              UF %in% Sudeste ~ "Sudeste",
                              UF %in% Sul ~ "Sul")))
```

The graph below shows the rural poverty rate values for the Brazilian states.

```{r, echo = TRUE}
fgt0_uf_rural %>% 
    dplyr::as_tibble() %>%
    dplyr::mutate(UF = forcats::fct_reorder(UF, VD5008)) %>%
    ggplot2::ggplot(ggplot2::aes(x = VD5008, y = UF)) +
    ggplot2::geom_col(aes(fill = regiao)) +
    ggplot2::labs(title = "Poverty rate by UF - Rural areas (2019)",
                  subtitle = "Considering the US$ 5.50 PPP line.",
                  x = NULL,
                  y = NULL,
                  caption = "Source: IBGE/PNADC microdata.") +
    ggplot2::geom_text(aes(label= round(VD5008,2)), vjust=0.5, hjust=-0.2, size=2.5)
```



